{\bf Intrusion Detection using Parzen-Windows on Provenance Graph Statistics}

### 2013-12-10 writeup/paper-final.tex
ran text through word to catch grammar mistakes. found a handful.

More recent work make use of the structure of provenance as a directed graph for a semi-automatic form
of intrusion detection, and to determine the scope of an intrusion. King et al. developed Backtracker \cite{backtracker}
 a modified
Linux kernel that tracks dependencies between operating system objects (files, processes, file names) similar
to provenance graphs. Backtracker builds a backwards causal graph to determine the entry point of an intrusion,
and it builds a forward causal graph to determine possibly tainted files, processes, or hosts in a distributed
system. However, Backtracker does not provide a way of automatically determining if an intrusion occurred
using provenance data; the provenance graph structure is only utilized after-the-fact.

@@

\subsubsection{hello}
We found that many of the popular exploits \cite{metasploit, exploitdb} involved buffer overflows and arbitrary code execution to possibly obtain root access. We used this as an guide to create a simple exploited \texttt{hello} program that echoes standard in to standard out, and we gave it a buffer overflow vulnerability that allows a malicious user to fork a shell or other process.

@@

\begin{figure*}
  \label{hello-orbiter}
  \caption{Descendants of a \texttt{hello} program. The top image shows multiple invocations of a \texttt{hello} behaving normally when invoked with \texttt{cat foo.txt | ./hello > c.txt}. The bottom image shows \texttt{hello} being exploited to fork an \texttt{ls} process. The gray circles are a provenance representation of a ``pipe''} 
  \centering
    \includegraphics[width=\textwidth]{img/hello.png}
\end{figure*}
\begin{figure*}
  \label{mcrypt-orbiter}
  \caption{Descendants of an unexploited mcrypt-style program. Simply an output file. Below, descendants of an exploited mcrypt-style program. In addition to an (offscreen) output file, there is also an outgoing \texttt{/bin/sh} process, with its own children.}
  \centering
    \includegraphics[width=\textwidth]{img/mcrypt.png}
\end{figure*}

@@

In Table 1, we show a summary of the statistics that had any statistical difference between normal and abnormally behaving nodes, and Figure 4 highlights the most salient statistics for the \texttt{gcc} exploit. The buffer overflows showed the most abnormal behavior in the number of outgoing processes and the centrality measures. This makes sense, as arbitrary code execution will generate additional processes. Of the centrality metrics we implemented, the Opsahl Centrality metric did the best in distinguishing an exploited node from normally behaving nodes. Because the Opsahl Centrality metric accounts for the number of edges between a node and all of its descendants, this centrality metric is very sensitive to the addition of nodes. Exploits that have additional outputs (e.g. new processes and output files through a remote shell), thus generate a higher Opsahl centrality metric for the offending node. The more active the exploit is (say, if it forks a user shell), the more the centrality measure will be impacted, and the more likely it is that the exploit will be caught by our model.

@@

We believe that the reason we had more promising results on our simulated intrusion data sets from PASS was because the provenance graphs from these workloads were much smaller (on the order of 5000 nodes). As a result of the much smaller size, small changes in provenance graph structure from an intrusion has a much profounder impact on the smaller provenance graphs. Thus, intrusions on our PASS workload had a much higher impact on the centrality calculations and resulted in more anomalous behavior in our Parzen-Window model. In the DARPA data set, we faced much larger provenance graphs (on the order of 20,000 nodes), and because of the much larger size, the effect of small changes in provenance graph structure from an intrusion had a much smaller impact on the overall structure of the graph. Intrusions did not change the centrality metrics very much in large graphs; therefore anomalous behavior went undetected, regardless of the window size we chose. Since our system could not distinguish between anomalous and normal behavior, then the true positive rate and false positive rate are essentially identical. In addition, we found that many more nodes were being flagged as intrusions (indeed, a positive result meant that a majority of nodes in a given time range were considered anomalous). In Figure 7, we see that the percentage of nodes during exploit windows were identified similarly to those during windows containing no exploit. As expected, as PSI increases, so too does the number of events marked as exploits. This result differed greatly from our preliminary exploration, where normal nodes were consistently identified as normal. We believe this occurred because the DARPA dataset came from a real system running for days, so the variety of nodes increased, and the distribution of these metrics became more full, so complete outliers were less present. 
\begin{figure}
  \label{intrusion-v-psi}
  \centering
    \includegraphics[width=0.5\textwidth]{img/intrusion_v_psi.png}
    \caption{The percentage of nodes flagged as an intrusion for various values of PSI. This figure was generated from the Week 5 Monday data, using a window size of 10 seconds. Other window sizes showed similar results. Recall that our system flags events with a percentage of 0.5 or greater as ``anomalous''} 
\end{figure}
From these results, we propose the following requirements for intrusion detection models that use provenance graph statistics in order to be useful for large provenance graphs:

@@

\subsection{N-distance Statistics}
Some processes fork helper processes that do the real work (such as \texttt{gcc}). We found that \texttt{gcc} forks processes for the linker/assembler that handles generating output, so \texttt{gcc} does not directly have output files. Similarly, when running a Python script, the code file itself is not directly executed, it is read in by a \texttt{python} process, which then goes off and collects input, generates output, and forks additional processes. Thus, when running a Python file, the file itself only has one node in the provenance graph, with many \texttt{python} process nodes accessing it. In order to account for these kinds of relationships, we could calculate the relevant statistics for the test node and all nodes at distance $N$ from the test node, and then aggregate them. By utilizing information about neighboring nodes, we hope to detect intrusions that involve more complicated relationships
between nodes.
\subsection{ENV and ARGV}
Currently we ignore the environment and argument information stored in the provenance graph. By generating histograms based on this information, we could catch subtler intrusions that rely on modified environment information. Additionally, if an exploit relies on a particular argument or argument patterns, this information might reflected in the ARGV information (consider a program which does not limit the number or size of command line arguments, and uses this vulnerability to execute malicious code).
\subsection{Different values for the density variance}
In our system, we used the variance $D$ of each data set as the variance of our kernel. To possibly get better results, one might try using different values for the variance, possible values that capture more granularity of students.
\subsection{Online Intrusion Detection}
Currently, this system runs after-the-fact, but truly useful intrusion detection systems need to run while the system is online. To do this, we would need to modify our statistics to use centrality metrics that are {\em progressive}, i.e., they can be recomputed online without needing to recompute the metric on the entire graph.

>>>>

More recent works make use of the structure of provenance as a directed graph for a semi-automatic form
of intrusion detection, and to determine the scope of an intrusion. King et al. developed Backtracker \cite{backtracker}
 a modified
Linux kernel that tracks dependencies between operating system objects (files, processes, file names) similar
to provenance graphs. Backtracker builds a backwards causal graph to determine the entry point of an intrusion,
and it builds a forward causal graph to determine possibly tainted files, processes, or hosts in a distributed
system. However, Backtracker does not provide a way of automatically determining if an intrusion occurred
using provenance data; the provenance graph structure is only utilized after-the-fact.

@@

\subsubsection{hello}
We found that many of the popular exploits \cite{metasploit, exploitdb} involved buffer overflows and arbitrary code execution to possibly obtain root access. We used this as a guide to create a simple exploited \texttt{hello} program that echoes standard in to standard out, and we gave it a buffer overflow vulnerability that allows a malicious user to fork a shell or other process.

@@

\begin{figure*}
  \label{hello-orbiter}
  \caption{Descendants of a \texttt{hello} program. The top image shows multiple invocations of a \texttt{hello} behaving normally when invoked with \texttt{cat foo.txt | ./hello > c.txt}. The bottom image shows \texttt{hello} being exploited to fork an \texttt{ls} process. The gray circles are a provenance representation of a ``pipe''.} 
  \centering
    \includegraphics[width=\textwidth]{img/hello.png}
\end{figure*}
\begin{figure*}
  \label{mcrypt-orbiter}
  \caption{Descendants of an unexploited mcrypt-style program - simply an output file. Below, descendants of an exploited mcrypt-style program. In addition to an (offscreen) output file, there is also an outgoing \texttt{/bin/sh} process, with its own children.}
  \centering
    \includegraphics[width=\textwidth]{img/mcrypt.png}
\end{figure*}

@@

In Table 1, we show a summary of the statistics that had any statistical difference between normal and abnormally behaving nodes, and Figure 4 highlights the most salient statistics for the \texttt{gcc} exploit. The buffer overflows showed the most abnormal behavior in the number of outgoing processes and the centrality measures. This makes sense, as arbitrary code execution will generate additional processes. Of the centrality metrics we implemented, the Opsahl Centrality metric did the best in distinguishing an exploited node from normally behaving nodes. Because the Opsahl Centrality metric accounts for the number of edges between a node and all of its descendants, this centrality metric is very sensitive to the addition of nodes. Exploits that have additional outputs (e.g. new processes and output files through a remote shell), thus generate a higher Opsahl centrality metric for the offending node. The more active the exploit is (say, if it forks a user shell), the more the centrality measure will be impacted, and the more likely it is that our model will catch the exploit.

@@

We believe that the reason we had more promising results on our simulated intrusion data sets from PASS was because the provenance graphs from these workloads were much smaller (on the order of 5000 nodes). As a result of the much smaller size, small changes in provenance graph structure from an intrusion has a much profounder impact on the smaller provenance graphs. Thus, intrusions on our PASS workload had a much higher impact on the centrality calculations and resulted in more anomalous behavior in our Parzen-Window model. In the DARPA data set, we faced much larger provenance graphs (on the order of 20,000 nodes), and because of the much larger size, the effect of small changes in provenance graph structure from an intrusion had a much smaller impact on the overall structure of the graph. Intrusions did not change the centrality metrics very much in large graphs; therefore anomalous behavior went undetected, regardless of the window size we chose. Since our system could not distinguish between anomalous and normal behavior, then the true positive rate and false positive rate are essentially identical. In addition, we found that many more nodes were being flagged as intrusions (indeed, a positive result meant that a majority of nodes in a given time range were considered anomalous). In Figure 7, we see that the percentage of nodes during exploit windows were identified similarly to those during windows containing no exploit. As expected, as PSI increases, so too does the number of events marked as exploits. This result differed greatly from our preliminary exploration, where normal nodes were consistently identified as normal. We believe this occurred because the DARPA dataset came from a real system running for days, so the variety of nodes increased, and the distribution of these metrics became more full, so complete outliers were less present. 
\begin{figure}
  \label{intrusion-v-psi}
  \centering
    \includegraphics[width=0.5\textwidth]{img/intrusion_v_psi.png}
    \caption{The percentage of nodes flagged as an intrusion for various values of PSI. This figure was generated from the Week 5 Monday data, using a window size of 10 seconds. Other window sizes showed similar results. Recall that our system flags events with a percentage of 0.5 or greater as ``anomalous''.} 
\end{figure}
From these results, we propose the following requirements for intrusion detection models that use provenance graph statistics in order to be useful for large provenance graphs:

@@

\subsection{N-distance Statistics}
Some processes fork helper processes that do the real work (such as \texttt{gcc}). We found that \texttt{gcc} forks processes for the linker/assembler that handles generating output, so \texttt{gcc} does not directly have output files. Similarly, when running a Python script, the code file itself is not directly executed, it is read in by a \texttt{python} process, which then goes off and collects input, generates output, and forks additional processes. Thus, when running a Python file, the file itself only has one node in the provenance graph, with many \texttt{python} process nodes accessing it. In order to account for these kinds of relationships, we could calculate the relevant statistics for the test node and all nodes at distance $N$ from the test node, and then aggregate them. By utilizing information about neighboring nodes, we hope to detect intrusions that involve more complicated relationships
between nodes.
\subsection{ENV and ARGV}
Currently we ignore the environment and argument information stored in the provenance graph. By generating histograms based on this information, we could catch subtler intrusions that rely on modified environment information. Additionally, if an exploit relies on a particular argument or argument patterns, this information might be reflected in the ARGV information (consider a program which does not limit the number or size of command line arguments, and uses this vulnerability to execute malicious code).
\subsection{Different values for the density variance}
In our system, we used the variance $D$ of each data set as the variance of our kernel. To possibly get better results, one might try using different values for the variance, possible values that capture more granularity of students.
\subsection{Online Intrusion Detection}
Currently, this system runs after-the-fact, but truly useful intrusion detection systems need to run while the system is online. To do this, we would need to modify our statistics to use centrality metrics that are {\em progressive}, i.e., they can be recomputed online without needing to recompute the metric on the entire graph.
